---
title: "Differential Marker Expression Analysis of Platelet Data"
author: "Arend Lis, Bernett Judith, Manz Quirin "
date: "1/18/2021"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(CATALYST)
library(flowCore)
library(ggplot2)
library(SingleCellExperiment)
library(data.table)
library(readxl)
library(diffcyt)
library(RColorBrewer)
library(grid)
library(futile.logger)
library(VennDiagram)
```

Data from human platelets of patients with chronic coronary syndrome undergoing different therapy: dual antiplatelet therapy versus triple antiplatelet therapy, before and after platelet activation with 10µm TRAP. There are files of 7 patients with triple therapy and 12 patients with dual therapy (each in two condtions)."

# Data Upload and Conversion to Single Cell Experiment
You can create a Single Cell Experiment of the fcs files using this code. You just have to specify the path to your data folder. 
```{r,eval=FALSE}
path <- "/Users/lisiarend/Desktop/Uni/Master/SysBioMed/CyTOF/extdata/HumanPlatelets"
exp <-list.files(path,pattern = "\\.fcs$",full.names = T)
names <- list.files(path,pattern = "*.fcs")
samples <-sapply(names, function(x) {
    strsplit(x, split = ".fcs")[[1]]
  })
patients <-sapply(names, function(x) {
    str_sub(strsplit(x, split = "_")[[1]][1], end = -2)
  })
a_b <- sapply(names, function(x) {
    str_sub(strsplit(x, split = "_")[[1]][1], start = 8)
  })
dual_triple <-
  sapply(names, function(x) {
    strsplit(strsplit(x, split = "_")[[1]][2], split = ".fcs")[[1]][1]
  })

metadata <- data.table(
  "file_name" = names,
  "sample_id" = samples,
  "patient_id" = patients,
  "activated_baseline" = a_b,
  "dual_triple" = dual_triple
)

panel <-read_excel(paste0(path,"panel_platelets.xlsx"))

sce <-
  prepData(
    exp,
    panel,
    metadata,
    transform = TRUE,
    cofactor = 5,
    md_cols = list(
      file = "file_name",
      id = "sample_id",
      factors = c("activated_baseline", "dual_triple","patient_id")
    )
  )
```

Otherwise, you can load the "sce_transformed.rds" file which already contains the SingleCellExperiment.
```{r}
path_to_rds <- "/Users/lisiarend/Desktop/Uni/Master/SysBioMed/CyTOF/cytof/data/platelets/sce_transformed.rds"
sce <- readRDS(path_to_rds)
```


# Data Preprocessing





# Data Visualization




# Clustering




# Differential Expression Analysis
