---
title: "Testing methods"
output: html_notebook
---

This notebook is for testing methods on the semi-simulated dataset of the diffcyt package. There are three kinds of datasets: 
1. Main dataset: contains spiked in B-cells 
2. Less distinct dataset 50%
3. Less distinct dataset 75%

Skip step 0 if you already have the SCE objects
## 0. Make SCE objects

The .fcs files have already been generated, now we have to read them in as a SCE object

```{r}
library(CATALYST)
library(data.table)
library(diffcyt)
library(flowCore)
library(SummarizedExperiment)
library(magrittr)
```

Make a function for the SCE objects: 

```{r}
produceSCE <- function(flowSet, metadata, panel, pop_names, pop_assignment){
  cn <- colnames(flowSet@frames[[metadata$file_name[1]]])
  #pretend that population is a marker
  cn[cn == "population"] <- pop_assignment
  for(elem in metadata$file_name){
    colnames(flowSet@frames[[elem]]@exprs) <- cn
  }
  colnames(flowSet) <- cn
  
  sce <- prepData(flowSet, panel = panel, md = metadata)
  mapping <- setNames(pop_names$name, pop_names$label)
  colData(sce)$cluster_id <- as.factor(mapping[assays(sce)$counts["population", ]])
  metadata(sce)$cluster_codes <- data.frame(population = as.factor(levels(colData(sce)$cluster_id)))

  #delete population from assays
  sce <- filterSCE(sce, !(marker_name %in% "population"))
  return(sce)
}
```


Main dataset: 

```{r}
files <- list.files("./benchmark_data/BCR_XL_sim/data/main/", pattern = "\\.fcs", full.names = T)
flowSetMain <- read.flowSet(files, transformation = F, truncate_max_range = F)
metadata <- fread("./benchmark_data/BCR_XL_sim/data/main/metadata.csv")
panel <- fread("./benchmark_data/BCR_XL_sim/data/main/panel.csv")
panel <- rbind(panel, data.table(fcs_colname = "population(Population185)Dd", antigen = "population", marker_class = "type"))
pop_names <-  fread("benchmark_data/BCR_XL_sim/data/population_names/population_names.csv")

sceDiffcytMain <- produceSCE(
  flowSet = flowSetMain, metadata = metadata, panel = panel, pop_names = pop_names, pop_assignment = "population(Population185)Dd"
  )

saveRDS(sceDiffcytMain, "./data/diffcyt/main/sce.rds")
saveRDS(panel, "./data/diffcyt/main/panel.rds")
saveRDS(metadata, "./data/diffcyt/main/md.rds")
```

Now do the same thing for the less distinct data set: 

First make a function for the metadata
```{r}
makeMetadata <- function(filenames){
  patientIDs <- unlist(sapply(filenames, tstrsplit, "\\_", keep=4))
  condition <- unlist(sapply(filenames, tstrsplit, "\\_", keep=5))

  metadata <- data.table(
    file_name = filenames,
    condition = condition,
    patient_id = patientIDs
  )
  metadata <- metadata %>% tidyr::unite("sample_id", condition:patient_id, sep = "_", remove = F)
}
```


1. less_50pc

```{r}
files <- list.files("./benchmark_data/BCR_XL_sim/data/less_distinct/less_50pc/", pattern = "\\.fcs", full.names = T)
flowSet50pc <- read.flowSet(files, transformation = F, truncate_max_range = F)
metadata50pc <- makeMetadata(list.files("./benchmark_data/BCR_XL_sim/data/less_distinct/less_50pc/", pattern = "\\.fcs"))
#use same panel
sceDiffcyt50pc <- produceSCE(
  flowSet = flowSet50pc, metadata = metadata50pc, panel = panel, pop_names = pop_names, pop_assignment = "population(Population185)Dd"
  )
saveRDS(sceDiffcyt50pc, "./data/diffcyt/less_50pc/sce.rds")
saveRDS(panel, "./data/diffcyt/less_50pc/panel.rds")
saveRDS(metadata50pc, "./data/diffcyt/less_50pc/md.rds")
```

2. less_75pc

```{r}
files <- list.files("./benchmark_data/BCR_XL_sim/data/less_distinct/less_75pc/", pattern = "\\.fcs", full.names = T)
flowSet75pc <- read.flowSet(files, transformation = F, truncate_max_range = F)
metadata75pc <- makeMetadata(list.files("./benchmark_data/BCR_XL_sim/data/less_distinct/less_75pc/", pattern = "\\.fcs"))
#use same panel
sceDiffcyt75pc <- produceSCE(
  flowSet = flowSet75pc, metadata = metadata75pc, panel = panel, pop_names = pop_names, pop_assignment = "population(Population185)Dd"
  )
saveRDS(sceDiffcyt75pc, "./data/diffcyt/less_75pc/sce.rds")
saveRDS(panel, "./data/diffcyt/less_75pc/panel.rds")
saveRDS(metadata75pc, "./data/diffcyt/less_75pc/md.rds")
```

## 1. Read in SCE objects 

```{r}
pathToData = "./data/diffcyt"
sceDiffcytMain <- readRDS(paste0(pathToData, "/main/sce.rds"))
sceDiffcyt75pc <- readRDS(paste0(pathToData, "/less_75pc/sce.rds"))  
sceDiffcyt50pc <- readRDS(paste0(pathToData, "/less_50pc/sce.rds"))
```


## 2. Start Analysis

### 2.1. Main

```{r}
designMain <- createDesignMatrix(
  ei(sceDiffcytMain),
  cols_design = c("condition", "patient_id")
)

contrastMainLimma <- createContrast(
  c(0, 1, rep(0, 7))
)

contrastMainLMM <- createContrast(
  c(0, 1)
)

formulaMain <- createFormula(
  ei(sceDiffcytMain),
  cols_fixed = "condition",
  cols_random = "patient_id"
)

source("pipeline/shiny_functions.R")
markersToTest <- getMarkersToTest(sceDiffcytMain, "limma", "all")
```

```{r}
#limma 

outLimma <- diffcyt(
  d_input = sceDiffcytMain, 
  design = designMain,
  contrast = contrastMainLimma,
  analysis_type = "DS",
  method_DS = "diffcyt-DS-limma",
  clustering_to_use = "population",
  markers_to_test = markersToTest
)

toptableLimma <- as.data.table(diffcyt::topTable(outLimma$res,all=TRUE,format_vals=TRUE))
toptableLimma
```

```{r}
#LMM

outLMM <- diffcyt(
  d_input = sceDiffcytMain,
  formula = formulaMain,
  contrast = contrastMainLMM,
  analysis_type = "DS",
  method_DS = "diffcyt-DS-LMM",
  clustering_to_use = "population",
  markers_to_test = markersToTest
)

toptableLMM <- as.data.table(diffcyt::topTable(outLMM$res,all=TRUE,format_vals=TRUE))
toptableLMM
```

```{r}
source("server/cytoGLMM_functions.R")
out <- runCytoGLMM(
  sceDiffcytMain, 
  condition = "condition",
  group = "patient_id"
)
summary(out)
```




