---
title: "Differential Marker Expression Analysis"
author: "Arend Lis, Bernett Judith, Manz Quirin "
date: "12.01.2021"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=12, fig.height=8)
library(CATALYST)
library(flowCore)
library(ggplot2)
library(SingleCellExperiment)
library(data.table)
library(readxl)
library(diffcyt)
library(RColorBrewer)
library(grid)
library(futile.logger)
library(VennDiagram)
set.seed(1234)

source("shiny_functions.R")
source("../server/clusterFun.R")
```

# Upload Data, Preprocessing and Clustering

If you don't want to perform the whole subselection of patients, clustering an visualization methods, you can simply skip those steps and load the data and move on to the Comparison of Differential Marker Expression Analysis Methods.
```{r}
#path_to_rds <- "../data/platelets/sceSampledClustered.rds"
#sce <- readRDS(path_to_rds)
```

Otherwise, upload data here.
```{r, data}
path_to_rds <- "../data/platelets/sce_transformed.rds"
sce <- readRDS(path_to_rds)
```

Make patient selection based on preprocessing plots and clustering results
```{r, selection}
sce <- makePatientSelection(sce, c("RPS 124", "RPS 108", "RPS 96", "RPS 149"))
```

Sample data to 4 patients receiving dual treatment and 4 patients receiving triple treatment
```{r, samplingPatients, eval=FALSE}
ei <- metadata(sce)$experiment_info
dual <- unique(ei[ei$dual_triple == "dual",]$patient_id)
triple <- unique(ei[ei$dual_triple == "triple",]$patient_id)

set.seed(1234)
sampled_duals <- as.character(dual[sample(length(dual), 4, replace=FALSE)])
set.seed(1234)
sampled_triples <- as.character(triple[sample(length(triple), 4, replace=FALSE)])

sampled_patients <- c(sampled_triples, sampled_duals)
sampled_patients

# sampled patients should be "RPS 136" "RPS 099" "RPS 127" "RPS 098" "RPS 154" "RPS 137" "RPS 139" "RPS 103"

sce <- filterSCE(sce, patient_id %in% sampled_patients)
```

Sample data to 20 000 cells for each sample
```{r, samplingEvents}

ei <- metadata(sce)$experiment_info
sample_ids <- ei$sample_id
coldata <- data.table()

coldata <- colData(sce)
rownames(coldata) <- seq(1:nrow(coldata))

sampled_indices <- c() 
for (id in sample_ids){
  dt <- coldata[coldata$sample_id == id,]
  indices <- rownames(dt)
  set.seed(1234)
  sample <- as.numeric(indices[sample(length(indices), 20000, replace=FALSE)])
  sampled_indices <- c(sampled_indices, sample)
}

sampled_indices <- sort(sampled_indices)

sce <- sce[,sampled_indices]

# metadata experiment_info does not update when subselecting
ei$n_cells <- rep(20000, nrow(ei))
metadata(sce)$experiment_info <- ei
```

Cluster the filtered SingleCellExperiment and run dimensionality reduction methods:
```{r}
sce <- clusterSCE(sce) 

DR_methods <- c("UMAP", "TSNE", "PCA", "MDS", "DiffusionMap", "Isomap")
for( method in DR_methods ){
  print(method)
  sce <- runDimRed(sce, method, cells_chosen = 100, feature_chosen = "type", assay_chosen = "exprs", scale = T, k = 3)
}

#saveRDS(sce,"../data/platelets/sceSampledClustered.rds")

```


# Comparison of Differential Marker Expression Analysis Methods

```{r, markerSpecification, include=FALSE}
is_marker <- rowData(sce)$marker_class %in% c("type","state")
id_all_markers <- (rowData(sce)$marker_class %in% c("type","state"))[is_marker] # type and state (but not none)
#id_type_markers <- (rowData(sce_dual_ab)$marker_class == "type")[is_marker] # type markers
#id_state_markers <- (rowData(sce_dual_ab)$marker_class == "state")[is_marker] # state markers
```

## Dual Activated vs. Dual Baseline

First, make the subselection of dual patients to compare activated and baseline platelets.
```{r, dualAB_selection}
allsamples <- unique(colData(sce)$sample_id)
dual_ab <- grep("dual",allsamples,value=TRUE)
sce_dual_ab <- filterSCE(sce, sample_id %in% dual_ab) # not using makeSampleSelection dual_ab are the including samples
```

Running diffcyt methods:
```{r, diffcyt}
parameters <- prepDiffExp(sce_dual_ab, contrastVars = c("activated_baseline"), colsDesign = c("activated_baseline"), method = "diffcyt-DS-limma")

#blockID <- metadata(sce_dual_ab)$experiment_info[["patient_id"]]

limma_dual_ab <- diffcyt::diffcyt(d_input = sce_dual_ab,
                 design = parameters[["design"]],
                 contrast = parameters[["contrast"]],
                 analysis_type = "DS",
                 method_DS = "diffcyt-DS-limma",
                 clustering_to_use = "all",
                 markers_to_test = id_all_markers
                 #block_id = blockID
                 )


parameters <-  prepDiffExp(sce_dual_ab, contrastVars = c("activated_baseline"), colsFixed = c("activated_baseline"), colsRandom=c("patient_id"), method = "diffcyt-DS-LMM")


LMM_dual_ab <- diffcyt::diffcyt(d_input = sce_dual_ab,
                 formula = parameters[["formula"]],
                 contrast = parameters[["contrast"]],
                 analysis_type = "DS",
                 method_DS = "diffcyt-DS-LMM",
                 clustering_to_use = "all",
                 markers_to_test = id_all_markers)

```

Let's try SigEMD:
```{r, eval=FALSE}
library(BiocParallel)
# Set the parameters and register the back-end to be used
param <- MulticoreParam(workers = 22)
register(param)

timed1 <- system.time(emd_result <- SigEMD(sce[, colData(sce)$dual_triple == "dual"], "all", "activated_baseline", nperm=100, parallel=TRUE))
#     user   system  elapsed 
# 4346.330 1598.441 2182.108 
print(timed1)

emd_result <- emd_result$all
source("../SigEMD/plot_sig.R")
lapply(emd_result$nonHur_gene, function(marker) plot_emd_density_sig(emd_result, marker))
```


```{r, eval=FALSE}
# now check if what happens if we skip the logistic regression
timed <- system.time(emd_result_allHur <- SigEMD(sce[, colData(sce)$dual_triple == "dual"], "all", "activated_baseline", Hur_gene=rownames(sce), nperm=100, parallel=TRUE, verbose = TRUE))
print(timed)
emd_result_allHur <- emd_result_allHur$all
lapply(state_markers(sce), function(marker) plot_emd_density_sig(emd_result, marker))
```


```{r, eval=FALSE}
# How much was contributed from the zeros?
print(emd_result$emdall[emd_result$nonHur_gene, ])
print(emd_result_allHur$emdall[emd_result$nonHur_gene, ])
```

Let's try out DEsingle:

```{r, eval=FALSE}
library(zinbwave)

group <- colData(sce)[["dual_triple"]]

# Detecting the DE genes in parallelization with 22 cores
results <- DEsingle(counts = sce, group = group, parallel = TRUE, BPPARAM = param)

# Dividing the DE genes into 3 categories at threshold of FDR < 0.05
results.classified <- DEtype(results = results, threshold = 0.05)
```

Significant markers found by all methods:

```{r  dual_ab_venn}
sign_limma <- data.frame(rowData(limma_dual_ab$res))[c("cluster_id","marker_id","p_val","p_adj")]
sign_LMM <- data.frame(rowData(LMM_dual_ab$res))[c("cluster_id","marker_id","p_val","p_adj")]
sign_limma$significant <- sign_limma$p_adj < 0.05
sign_LMM$significant <- sign_LMM$p_adj < 0.05

limma <- unlist(subset(sign_limma, significant==TRUE, select=c(marker_id), use.names=FALSE))
LMM <- unlist(subset(sign_LMM, significant==TRUE, select=c(marker_id), use.names=FALSE))


## include SigEMD and ZINB

library(ggvenn)
  
venn <- venn.diagram(
  x = list(limma,LMM),
  filename = "../plots/dual_ab_venn.png",
  category.names = c("limma","LMM"),
  image.type = "png",
  cex = 2.0,
  fontfamily = "sans",
  col=c("#99CC99", "#CC3333"),
  #fill = c(alpha("#99CC99",0.3), alpha("#CC3333",0.3)),
  cat.cex = 1.8,
  cat.fontface = "bold",
  cat.fontfamily = "sans",
  main = paste("Limma vs. LMM"),
  main.fontface = "bold",
  main.fontfamily = "sans",
  main.cex = 2.0,
  sub = paste("P-adjusted < 0.05"),
  sub.cex=1.5 ,
  sub.fontfamily="sans"
  )

```

```{r}
sce_small <- readRDS("../data/platelets_small/sce.rds")
is_marker <- rowData(sce_small)$marker_class %in% c("type","state")
id_all_markers <- (rowData(sce_small)$marker_class %in% c("type","state"))[is_marker] # type and state (but not none)
allsamples <- unique(colData(sce_small)$sample_id)
dual_ab <- grep("dual",allsamples,value=TRUE)
sce_small_dual_ab <- filterSCE(sce_small, sample_id %in% dual_ab) # not using makeSampleSelection dual_ab are the including samples
parameters <-  prepDiffExp(sce = sce_small_dual_ab, contrastVars = c("activated_baseline"), colsFixed = c("activated_baseline"), colsRandom=c("patient_id"), method = "diffcyt-DS-LMM")
LMM_dual_ab <- diffcyt::diffcyt(d_input = sce_small_dual_ab,
                 formula = parameters[["formula"]],
                 contrast = parameters[["contrast"]],
                 analysis_type = "DS",
                 method_DS = "diffcyt-DS-LMM",
                 clustering_to_use = "all",
                 markers_to_test = id_all_markers)
data.table::as.data.table(diffcyt::topTable(LMM_dual_ab$res, all=TRUE))
sign_LMM <- data.frame(rowData(LMM_dual_ab$res))[c("cluster_id","marker_id","p_val","p_adj")]
sign_LMM$significant <- sign_LMM$p_adj < 0.05
LMM <- unlist(subset(sign_LMM, significant==TRUE, select=c(marker_id), use.names=FALSE))
CATALYST::plotDiffHeatmap(x = sce_small_dual_ab,
                          y = rowData(LMM_dual_ab$res), 
                          all = T)
```

## Triple Activated vs. Triple Baseline

## Dual Activated vs. Triple Activated

## Dual Baseline vs. Triple Basleline



