---
title: "Differential Marker Expression Analysis of Platelet Data"
author: "Arend Lis, Bernett Judith, Manz Quirin "
date: "1/18/2021"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=12, fig.height=8)
library(data.table)
library(CATALYST)
library(flowCore)
library(ggplot2)
library(SingleCellExperiment)
library(readxl)
library(diffcyt)
library(RColorBrewer)
library(grid)
library(futile.logger)
library(VennDiagram)
library(stringr)
set.seed(1234)

source("shiny_functions.R")
source("../server/clusterFun.R")
```

Data from human platelets of patients with chronic coronary syndrome undergoing different therapy: dual antiplatelet therapy versus triple antiplatelet therapy, before and after platelet activation with 10µm TRAP. There are files of 7 patients with triple therapy and 12 patients with dual therapy (each in two condtions).

# Data Upload and Data Preprocessing
Fcs files, as well as metadata and a panel file are given for this dataset and uploaded in this section. You just have to specify the path to your files. The Single Cell Experiment is then created using the prepData function of CATALYST. The function prepData can directly transform the marker intesities using arcsinh (inverse hyerpoblic sine) to make the distributions more symmetrix and to map them to a comparable range of expression. Recommended values for the cofactor parameter are 5 for mass cytometry (CyTOF) or 150 for fluorescence flow cytometry

```{r,eval=FALSE}
path <- "/Users/lisiarend/Desktop/Uni/Master/SysBioMed/CyTOF/extdata/HumanPlatelets"
#path <- "/Users/judith_bernett/Desktop/Bioinfo/SystemsBioMedicine/systems_biomedicine_spill_applied_fcs_files/"
cofactor <- 5
exp <-list.files(path,pattern = "\\.fcs$",full.names = T)#[1:14]
names <- list.files(path,pattern = "*.fcs")#[1:14]
samples <-sapply(names, function(x) {
    strsplit(x, split = ".fcs")[[1]]
  })
patients <-sapply(names, function(x) {
    str_sub(strsplit(x, split = "_")[[1]][1], end = -2)
  })
a_b <- sapply(names, function(x) {
    str_sub(strsplit(x, split = "_")[[1]][1], start = 8)
  })
dual_triple <-
  sapply(names, function(x) {
    strsplit(strsplit(x, split = "_")[[1]][2], split = ".fcs")[[1]][1]
  })
metadata <- data.table(
  "file_name" = names,
  "sample_id" = samples,
  "patient_id" = patients,
  "activated_baseline" = a_b,
  "dual_triple" = dual_triple
)
panel <-read_excel(paste0(path,"panel_platelets.xlsx"))
sce <-
  prepData(
    exp,
    panel,
    metadata,
    transform = TRUE,
    cofactor = cofactor,
    md_cols = list(
      file = "file_name",
      id = "sample_id",
      factors = c("activated_baseline", "dual_triple","patient_id")
    )
  )
```

Otherwise, you can load the "sce_transformed.rds" file which already contains the transformed SingleCellExperiment.

```{r}
path_to_rds <- "/nfs/home/students/l.arend/data/platelets/sce_transformed.rds"
sce <- readRDS(path_to_rds)

#or load already clustered and filtered SingleCellExperiment
#path_to_rds <- "../data/platelets/sceFilteredClustered.rds"
#sce <- readRDS(path_to_rds)
```


The shiny app also includes some simple visualization plots to verify whether the data represents what we expect, for example, whether samples that are replicates of one condition are more similar and are distinct from samples from another condition. Depending on the situation, one can then consider removing problematic markers or samples from further analysis.

```{r, fig.width=12, fig.height=8}
CATALYST::plotCounts(
    sce,
    group_by = "patient_id", # possible inputs: names(colData(sce))
    color_by = "dual_triple", # possible inputs: names(colData(sce))
    prop = FALSE # TRUE for stacked relative abundances, FALSE for total cell counts
  )

CATALYST::pbMDS(
      sce,
      label_by = "patient_id", 
      color_by = "dual_triple",
      features = "type", # possible inputs: unique(rowData(sce)$marker_class) or NULL (to use all features)
      assay = "exprs", # possible inputs: assayNames(sce)
    )

CATALYST::plotNRS(
      sce,
      color_by = "activated_baseline",
      features = "state",
      assay = "exprs"
    )

CATALYST::plotExprs(
      sce,
      color_by = "activated_baseline",
      features = "type",
      assay = "exprs"
    )

CATALYST::plotExprs(
      sce,
      color_by = "activated_baseline",
      features = "state",
      assay = "exprs"
    )

CATALYST::plotExprs(
      sce,
      color_by = "dual_triple",
      features = NULL,
      assay = "exprs"
    )

CATALYST::plotExprHeatmap(
      sce,
      features = NULL,
      assay = "exprs"
    )
```

These plots can then be used to select patients as well as samples and filter the Single Cell Experiment. As one can see, the number of cells in patient RPS124 is notably lower than in the other patients. Additionally, the MDS plot shows that patients RPS108 and RPS124 are separated from the others. The expression heatmap agrees with our observations. Patients RPS124 and RPS128 have a weird expression. Additionally, the sample RPS149_A shows a strange expression over all markers. A deeper look into these patients is taken after the clustering.

Furthermore, the state markers show strong differences between activated and baseline platelets. Comparing dual treatment versus triple treatment, no big difference in marker expression is noticeable.

# Clustering
Clustering is performed using Flowsom on the type markers.
```{r}
sce <- clusterSCE(sce)

plotly::ggplotly(delta_area(sce))
```

The delta area score (y-axis) indicates the relative increase in cluster stability obtained when clustering the 100 SOM codes generated by FlowSOM into k groups (x-axis). It allows a user to determine the relative increase in consensus and determine k at which there is no appreciable increase. We have chosen meta6 since there is no significant increase afterwards.

Let's have a look at the clusters:
```{r}
CATALYST::plotAbundances(sce, "meta6", group_by = "activated_baseline")

CATALYST::plotAbundances(sce, "meta6", by = "cluster_id", group_by = "activated_baseline")
```

```{r}
CATALYST::plotClusterExprs(sce, "meta6")

#CATALYST::plotFreqHeatmap(sce, "meta6")
```

As we can see above, some patients have odd clusters. Further investigation might be necessary to uncover the reasons for that. For now we will exclude the patients from the analysis. Hence, we will repeat the clustering step:
```{r}
sce <- makePatientSelection(sce = sce, deselected_patients = c("RPS 108", "RPS 096", "RPS 149", "RPS 124"))
sce <- clusterSCE(sce)

plotly::ggplotly(delta_area(sce))
```

The clustering show no significant increase after meta9. Let's have a look at the clusters:
```{r}
CATALYST::plotAbundances(sce, "meta6", group_by = "activated_baseline")

CATALYST::plotAbundances(sce, "meta6", by = "cluster_id", shape_by = "patient_id", group_by = "activated_baseline")
```

```{r, fig.width=6, fig.height=4}
invisible(plotStarsCustom(sce, backgroundValues = cluster_codes(sce)[["meta6"]]))
```

```{r}
CATALYST::plotClusterExprs(sce, "meta6")

CATALYST::plotFreqHeatmap(sce, "meta6")
```

Now, we can visualize the different clusters, but first we have to run the methods.

# Data Visualization
You can now run various dimensionality reduction methods on your data. The options are (parameter dr_chosen): 
<ul>
<li>"UMAP"</li>
<li>"TSNE"</li>
<li>"PCA"</li>
<li>"MDS"</li> 
<li>"DiffusionMap" and </li>
<li>"Isomap"</li>
</ul>
Other run options are:
<ul>
<li> cells_chosen: number of cells to sample from.  </li>
<li> feature_chosen: markers or marker class. We recommend using "type" </li>
<li>assay_chosen: "counts" or "exprs". We recommend using "exprs" </li> 
<li> scale: TRUE or FALSE. We recommend using TRUE </li>
<li> k: Isomap-specific parameter. Specifies the number of neighbours for the graph constructed by isomap </li>


First, we are performing a dimensionality reduction on the states markers for 300 cells and visualize the results.
```{r}
DR_methods <- c("UMAP","PCA")

for( method in DR_methods ){
  print(method)
  sce <- runDimRed(sce, method, cells_chosen = 3000, feature_chosen = "state", assay_chosen = "exprs", scale = T, k = 3)
}

for( method in DR_methods ){
  g <- CATALYST::plotDR(sce, dr = method, color_by = "activated_baseline", facet_by = "dual_triple")
  print(g)
}
```

Secondly, we are going to run the dimensionality reductions on the type markers and visualize the clusters.
```{r}
DR_methods <- c("UMAP","PCA")

for( method in DR_methods ){
  print(method)
  sce <- runDimRed(sce, method, cells_chosen = 1000, feature_chosen = "type", assay_chosen = "exprs", scale = T, k = 3)
}


for( method in DR_methods ){
  g <- CATALYST::plotDR(sce, dr = method, color_by = "meta6")
  print(g)
}
```

Let's check if we cluster by type, color by the expression of state markers, facetted by the conditions activated and baseline.
```{r}
DR_methods <- c("UMAP","PCA")

for( method in DR_methods ){
  print(method)
  sce <- runDimRed(sce, method, cells_chosen = 1000, feature_chosen = "type", assay_chosen = "exprs", scale = T, k = 3)
}

state_markers <- rowData(sce)[rowData(sce)$marker_class=="state",]$marker_name
for( method in DR_methods ){
  g <- CATALYST::plotDR(sce, dr = method, color_by = state_markers, facet_by="activated_baseline")
  print(g)
}
```

Finally, we are going to run the dimensionality reductions on all markers and show there is a difference between dual and triple.
```{r}
DR_methods <- c("UMAP","PCA")

for( method in DR_methods ){
  print(method)
  sce <- runDimRed(sce, method, cells_chosen = 1000, feature_chosen = NULL, assay_chosen = "exprs", scale = T, k = 3)
}

for( method in DR_methods ){
  g <- CATALYST::plotDR(sce, dr = method, color_by = "dual_triple", facet_by="activated_baseline")
  print(g)
}
```


After clustering and visualization, we can move to the differential expression analysis.

# Differential Expression Analysis

In the next step, 4 major comparisons are made. The dual activated conditions is compared to the dual baseline condition. This comparison is also performed for triple therapy (triple activated vs. triple baseline). In addition, dual activated is compared to triple activated and dual baseline is compared to triple baseline.

## Dual Activated vs. Dual Baseline
First, we have to filter the data to only dual patients.
```{r}
sce_d <- filterSCE(sce, dual_triple == "dual")

# ja oder nein?
rowData(sce_d)$marker_class <- rowData(sce_d)$marker_class[rowData(sce_d)$marker_class == "none"] <- "state"

```

Before performing the differential marker expression analysis, the median marker expressions are plotted.
```{r,fig.with=12, fig.height=8}

CATALYST::plotPbExprs(sce_d, k = "all", features = NULL , color_by = "activated_baseline", 
              facet_by = "antigen")

CATALYST::plotPbExprs(sce_d, k = "meta6", features = NULL , color_by = "activated_baseline", 
              facet_by = "cluster_id")

```

Finally, the differential analysis can be performed using the method limma.
```{r}
# subsett SingleCellExperiment for dual patients
res_d <- runDS(sce = sce_d,
      condition = "activated_baseline",
      de_methods = c("limma"),
      k = "all",
      features = "all",
      markers_to_test = "all"
      )
```

Now, we can visualize that: 
```{r,fig.with=12, fig.height=8}
# change ordering of sample_ids for heatmap
samples <- metadata(sce_d)$experiment_info$sample_id
activated <- grep("A", samples, value=TRUE)
baseline <- grep("B", samples, value=TRUE)
colData(sce_d)$sample_id <- factor(colData(sce_d)$sample_id, level=c(activated, baseline), labels=c(activated,baseline))

# levels für meta data auch ändern
CATALYST::plotDiffHeatmap(sce_d, rowData(res_d$limma$res), all=T, col_anno = "activated_baseline", normalize = FALSE)
```

## Triple Activated vs. Triple Baseline

```{r}
sce_t <- filterSCE(sce, dual_triple == "triple")
```

```{r,fig.with=12, fig.height=8}

CATALYST::plotPbExprs(sce_t, k = "all", features = NULL , color_by = "activated_baseline", 
              facet_by = "antigen")

CATALYST::plotPbExprs(sce_t, k = "meta6", features = NULL , color_by = "activated_baseline", 
              facet_by = "cluster_id")

```

```{r}
res <- runDS(sce = sce_t,
      condition = "activated_baseline",
      de_methods = c("limma"),
      k = "all",
      features = "all",
      )

```

```{r,fig.with=12, fig.height=8}
CATALYST::plotDiffHeatmap(sce_t, rowData(res$limma$res), all = T, normalize = FALSE)
```


## Dual Activated vs. Triple Activated
```{r}
sce_a <- filterSCE(sce, activated_baseline == "A")
```

```{r,fig.with=12, fig.height=8}

CATALYST::plotPbExprs(sce_a, k = "all", features = NULL , color_by = "dual_triple", 
              facet_by = "antigen")

CATALYST::plotPbExprs(sce_a, k = "meta6", features = NULL , color_by = "dual_triple", 
              facet_by = "cluster_id")

```

```{r}
res <- runDS(sce = sce_a,
      condition = "dual_triple",
      de_methods = c("limma"),
      k = "all",
      features = "all",
      )

```

```{r,fig.with=12, fig.height=8}
CATALYST::plotDiffHeatmap(sce_a, rowData(res$limma$res), all = T)
```

## Dual Baseline vs. Triple Basleline

```{r}
sce_b <- filterSCE(sce, activated_baseline == "B")
```

```{r,fig.with=12, fig.height=8}

CATALYST::plotPbExprs(sce_b, k = "all", features = NULL , color_by = "dual_triple", 
              facet_by = "antigen")

CATALYST::plotPbExprs(sce_b, k = "meta6", features = NULL , color_by = "dual_triple", 
              facet_by = "cluster_id")

```

```{r}
res <- runDS(sce = sce_b,
      condition = "dual_triple",
      de_methods = c("limma"),
      k = "all",
      features = "all",
      )

```

```{r,fig.with=12, fig.height=8}
CATALYST::plotDiffHeatmap(sce_b, rowData(res$limma$res), all = T)
```



